The error when inferring phylogenies with incipient species by a Birth-Death model

### 2018-04-12 article.tex
Note about Moeller et al., 2018 article

  % One or two sentences providing a basic
  % introduction to the field,
  % comprehensible to a scientist in any discipline.
  The tools for reconstructing phylogenetic relationships between taxonomic 
  units (e.g. species) have become very advanced in the last two decades.

@@

The computational tools a contemporary phylogeneticist has at his/her disposal
goes beyond the wildest imagination of those living two decades ago. 
Advances in computational power allowed the first cladograms to be inferred 
from DNA alignments in 1981 \cite{felsenstein1981}. The first Bayesian tools 
appeared in 1996 \cite{rannala1996}, allowing
for unprecedented flexibility in the setup of a phylogenetic model.

Currently, the most popular Bayesian phylogenetics tools are BEAST \cite{beast},  
MrBayes [ref] and RevBayes [ref]. Each of these use efficient tree sampling 
routines to create a posterior probability distribution of the to-be-inferred 
phylogenetic tree. But next to the computational speed, they also allow
to incorporate known or hypothesized structure of the phylogenetic 
tree-to-be-inferred through the model priors.

The model prior most relevant here, is the tree prior. The tree prior
embodies the assumed speciation model. Each speciation model has a
different expected branch length distribution. It has been shown 
[Moeller, du Plessis, Stadler. 2018] that the effect of the prior on the posterior distribution 
of trees can be substantial.

@@

From a species tree, we simulate a DNA alignment that has the same history
as the phylogeny. The nucleotides of the DNA alignment follow a JC69
nucleotide substitution model, in which all nucleotide-to-nucleotide transitions
are equally likely. Although this may seem as a simplification, in our Bayesian
inference (see below) we can use this exact site model as a prior, which we 
know is the truth.
The DNA sequence length and mutation rate was chosen as such to provide a
resolution of a 1000 years, that is, to have one expected nucleotide change 
per 1000 years per lineage on average. As our time unit is in 
millions of years, and our crown age is 15 million years,
this equates to a rate of $\frac{1000}{15}$ nucleotide substitutions 
changes per time unit per lineage. To provide for each substitution being unique,
we simulate as much DNA nucleotides as expected mutations, which equals the
nucleotide substitution rate times the crown age, which equals 15K. As this
mutation rate is constant over all branches, it in effect follows a strict 
clock model, which we will specify as the known clock model prior later.

>>>>

  % One or two sentences providing a basic
  % introduction to the field,
  % comprehensible to a scientist in any discipline.
  The tools for reconstructing phylogenetic relationships between taxonomic 
  units (e.g. species) have become very advanced in the last three decades.

@@

The computational tools a contemporary phylogeneticist has at his/her disposal
goes beyond the wildest imagination of those living three decades ago. 
Advances in computational power allowed the first cladograms to be inferred 
from DNA alignments in 1981 \cite{felsenstein1981}. The first Bayesian tools 
appeared in 1996 \cite{rannala1996}, allowing
for unprecedented flexibility in the setup of a phylogenetic model.

Currently, the most popular Bayesian phylogenetics tools are BEAST \cite{beast},  
MrBayes \cite{mrbayes} and RevBayes \cite{revbayes}. 
Each of these use efficient tree sampling 
routines to create a posterior probability distribution of the to-be-inferred 
phylogenetic tree. But next to the computational speed, they also allow
to incorporate known or hypothesized structure of the phylogenetic 
tree-to-be-inferred through the model priors.

In a Bayesian analysis, the model priors are explicitly specified.
Those model priors can be grouped in three groups: (1) site model, which
governs the nucleotide subsititution model, (2) clock model, specifying
the rate of mutation per lineage in time, and (3) tree prior, embodying
the speciation model behind branching events (speciation) 
and branch termination (extinction).

The effect of choosing a (potentially wrong) prior will have its effect on
the posterior. Recently, for example, \cite{moller2018} show that the effect
of choosing a tree prior biases the estimation of the molecular clock rate, 
for DNA sequences of 100-1000 base pairs.

This bias has its implications in the research and policy regarding rapidly 
evolving viruses. [NOTE: I would say that same study concludes 
that it does not matter for DNA alignments of 15K, which is the DNA 
alignment length that I use].

@@

From a species tree, we simulate a DNA alignment that has the same history
as the phylogeny. The nucleotides of the DNA alignment follow a JC69
nucleotide substitution model, in which all nucleotide-to-nucleotide transitions
are equally likely. Although this may seem as a simplification, in our Bayesian
inference (see below) we can use this exact site model as a prior, which we 
know is the truth.
The DNA sequence length and mutation rate was chosen as such to provide a
resolution of a 1000 years, that is, to have one expected nucleotide change 
per 1000 years per lineage on average. As our time unit is in 
millions of years, and our crown age is 15 million years,
this equates to a rate of $\frac{1000}{15}$ nucleotide substitutions 
changes per time unit per lineage. To provide for each substitution being unique,
we simulate as much DNA nucleotides as expected mutations, which equals the
nucleotide substitution rate times the crown age, which equals 15K (which is
coincidentally the same as \cite{moller2018}). As this
mutation rate is constant over all branches, it in effect follows a strict 
clock model, which we will specify as the known clock model prior later.
